ECD_Supp_Mat

### 2018-05-10 ECD_Supp_Mat.Rmd
updates to supp mat

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

$$\hat{R_i}=\frac{m}{e(a_{ii}+a_{aij})}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. An increased in total attack rate results in a decrease in resource abundances. Now let's assume that the evolution of consumer attack rates are constrained by a linear trade-off, i.e. a unit increase in $a_{ii}$ results in a unit decrease in $a_{ij}$. Although such a trade-off results in divergent character displacement (Abrams 1986), there is no net change in resource abundances when consumer's diverge. Although this equilibrium equation for resource abundances was derived for the scenario where both consumers and both resources are present, it predicts well the abundance of resources when there is only one consumer as well. This is likely because a consumer, in the absence of competition, evolves to be a generalist that has equal attack rates on each resources, resulting in equivalent resource abundances.

@@

```{r Trade-off Function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

@@

In contrast, the shape of the trade-off qualitatively affects the relationship between divergence and resource abundance. As mentioned previously, a linear trade-off results in no change in resource densities. If the trade-off is convex, then resource densities can actually increase under character displacement! This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{aij}$) and decreases as consumers diverge. Note though that detecting this effect in nature may be difficult, since a convex trade-off results in relatively small displacement relative to other trade-off forms (Lawlor and Smith 1976). When the trade-off is concave, however, we see that character divergence suppresses resource densities.

Most models of character displacement have assumed a scenario where consumers can forage for both resources simultaneously (i.e. resources occur in the same microhabitat). This assumption may be valid for some foraging scenarios, e.g. Darwin's finches foraging for seeds. However, in many situations, consumers forage for resources that occur in distinct microhabitats (McCann et al. 2005). The only chracter displacement model that we are aware of that modeled resources in distinct microhabitat was one examined by Lawlor and Smith (1976). They found that consumers still underwent character divergence, regardless of the foraging scenario. Interestingly, we find that the foraging scenario qualitatively affects the relationship between character divergence and resource abundances. Specifically, if consumers are competing for resources that occur in distinct microhabitats, then consumer divergence always results in resource suppression. This occurs because now the consumers effective attack rate is influenced by its habitat preference. Note also that this resource suppression occurs regardless of the shape of the trade-off, although the shape of the trade-off can amplify the effect of character divergence on resource supression. The type of trade-off does affect the amount of character displacement though, and actually, we see again the a concave trade-off results in less displacement and less of an effect on resource abundances; whereas a convex trade-off results in the largest displacement and the largest effect on resource abundances. This effect of spatial context on the relationship between character displacement and resource abundances appears to be quite general. For example, in the appendix, we analyze a model that was designed to capture the spatial coupling of consumers in many food-web context (McCann 2005). Here, we again find that divergence results in a negative effect on resource abundances.

We would like to note that the scenario where the relationship between character divergence and resource suppression is the strongest is when the trade-off in attack rates is concave and consumers are competiting for resources that occupy distinct habitat types. Given that prior work in threespine stickleback has shown a concave trade-off in feeding rates (Schluter 1993; Arnegard et al. 2014) and that stickleback ecotypes are competing for resources that occur in distinct microhabitats (benthic vs. limnetic), we predict that character divergence in sticklebacks would supress resource densities.

@@

## Testing the generality of our predictions

So far, our theoretical predictions of the effects of consumer divergence on resource dynamics has considered a simplistic scenario, where resources are equivalent and consumers have equivalent conversion efficiences, but have mirror image attack rates on each resource. We conjecture that these results are quite general, given that the process of character displacement is quite general. To explore the generality of our predictions, we examined whether our conclusions still held under a foraging scenario that was designed to mimic the stickleback lake food web. To do this, we used empirical relationships between body size and the parameters that govern consumer-resource dynamics to parameterize an eco-evolutionary model. Specifically, we used consumer and resource biomasses

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = TRUE,
	warning = TRUE,
	include = FALSE
)
source('eco_evo_simulations.R')
```

@@

$$\hat{R_i}=\frac{m}{e(a_{ii}+a_{aij})}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. An increased in total attack rate results in a decrease in resource abundances. Now let's assume that the evolution of consumer attack rates are constrained by a linear trade-off, i.e. a unit increase in $a_{ii}$ results in a unit decrease in $a_{ij}$. Although such a trade-off results in divergent character displacement (Abrams 1986), there is no net change in resource abundances when consumer's diverge.

## Trade-offs

@@

```{r Trade-off Function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
bind_rows(convex_df, linear_df, concave_df) %>%
  ggplot(., aes(x=aii, y=aij, color = factor(n))) + geom_line() + ggtitle("Attack rate trade-offs")
```

In contrast, the shape of the trade-off qualitatively affects the relationship between divergence and resource abundance. As mentioned previously, a linear trade-off results in no change in resource densities. If the trade-off is convex, then resource densities can actually increase under character displacement! This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{aij}$) and decreases as consumers diverge. Note though that detecting this effect in nature may be difficult, since a convex trade-off results in relatively small displacement relative to other trade-off forms (Lawlor and Smith 1976). When the trade-off is concave, however, we see that character divergence suppresses resource densities. Although this equilibrium equation for resource abundances was derived for the scenario where both consumers and both resources are present, it predicts well the abundance of resources when there is only one consumer as well. This is likely because a consumer, in the absence of competition, evolves to be a generalist that has equal attack rates on each resources, resulting in equivalent resource abundances.

```{r Effect of trade-off on ECD-Resource relationship, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

cbind.data.frame(tradeoff_df, MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))
```

## Foraging Scenario

Most models of character displacement have assumed a scenario where consumers can forage for both resources simultaneously (i.e. resources occur in the same microhabitat). This assumption may be valid for some foraging scenarios, e.g. Darwin's finches foraging for seeds. However, in many situations, consumers forage for resources that occur in distinct microhabitats (McCann et al. 2005). The only chracter displacement model that we are aware of that modeled resources in distinct microhabitat was one examined by Lawlor and Smith (1976). This model takes the same form as the MacArthur competition model, except now the consumer's functional response takes the form:

$$F_{ij}(R_j)=w_{ij}a_{ij}R_j$$

where $w_{ij}$ represents the proportion of time $C_i$ spends foraging in the habitat where $R_j$ is found. Note that since $w_ij$ is a proportion that $w_{ii}=1-w_{ij}$.

Lawlor and Smith found that consumers still underwent character divergence, regardless of the foraging scenario. Interestingly, we find that the foraging scenario qualitatively affects the relationship between character divergence and resource abundances. Note that now, equilibrium resource abundances are determined by the following equation:

$$\hat{R_i}=\frac{1}{w_{ii}a_{ii}+(1-w_{ii})a_{ij}}\cdot\frac{m}{e}$$

This means that if if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{ii}>0.5$ and $a_{ii}>a_{ij}$), then resource abundances will always decrease, regardless of the trade-off. Thus, divergent character displacement always results in resource suppression if consumers are competing for resources that occur in distinct microhabitats. Note that the shape of the trade-off can dampen or amplify the effect of character divergence on resource supression. This is because the shape of the trade-off influences the magnitude of character displacement. For example, a convex relationship results in less displacement and less of an effect on resource abundances; whereas a concave trade-off results in the largest displacement and the largest effect on resource abundances.

```{r}
ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))

cbind.data.frame(tradeoff_df, LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=Competitor),
             shape=21, size=10, inherit.aes = FALSE) +
  scale_fill_manual(values=c("white","black"))
```

This effect of spatial context on the relationship between character displacement and resource abundances appears to be quite general. For example, even if we consider a more realistic consumer functional response (derived by McCann et al. 2005):

$$F_{ii}(R_i,R_j)=\frac{a_{ii}W_{ii}R_i}{1+a_{ii}hW_{ii}R_i+a_{ij}hW_{ij}R_j}$$

where consumer feeding rates on $R_i$ are influenced by both resource densities; saturate as resource densities increase (due to handling time $h$); and consumer habitat preferences are modified by relative resource densities, such that:

$$W_{ii}=\frac{w_{ii}R_i}{w_{ii}R_i+(1-w_{ii})R_j}$$ 
we still observe the same qualitative relationship between character divergence and resource suppression. This is because resource competition always leads to character divergence (see proof in Appendix) and equilibrium resource abundances are governed by similar dynamics:

$$\hat{R_i}=\frac{1}{w_{ii}a_{ii}+(1-w_{ii})a_{ij}}\cdot\frac{m}{e-hm}$$

## Predictions for threespine stickleback

The scenario where the relationship between character divergence and resource suppression is the strongest is when the trade-off in attack rates is concave and consumers are competiting for resources that occupy distinct habitat types. Prior work in threespine stickleback suggests that the evolution of stickleback attack rates are constrained by a concave trade-off (Schluter 1993; Arnegard et al. 2014) and that stickleback ecotypes are competing for resources that occur in distinct microhabitats (benthic vs. limnetic). Thus, we predict that character divergence in sticklebacks would supress resource densities.

We note, however, that our analytical insights to the effects of consumer divergence on resource dynamics has considered a simplistic scenario (resources are equivalent and consumers have equivalent conversion efficiences, but have mirror image attack rates on each resource). To explore the generality of our predictions, we examined whether our conclusions still held under a foraging scenario that was designed to mimic the stickleback lake food web. To do this, we used empirical relationships between body size and the parameters that govern consumer-resource dynamics to parameterize an eco-evolutionary model. Specifically, we used consumer and resource biomasses

```{r Allometric Scaling of Stickleback Food Web}

mass_scaling <- function(W, intercept, exponent) {
  log_rate <- intercept + exponent * log(W)
  return(exp(log_rate))
}

## BODY SIZE (g) ----

g_zoop <- 0.004/1000            # covert 0.004 mg to g
g_benthic_invert <- 0.25/1000   # convert 0.25 mg to g

g_limnetic <- 0.81              # from Schluter 1993, Ecology Table 1
g_benthic <- 1.48               # from Schluter 1993, Ecology Table 1
g_solitary <- 1.12              # from Schluter 1993, Ecology Table 1

## INTRINSIC GROWTH RATE (r) ----
# r = birth rate per day
# proportional to M^(-1/4) from Rip and McCann 2011 (they cite Brown et al. 2004)
# alternatively use scaling exponent = -0.27377 for heterotherm metazoa from Fenchel 1974, Oecologia
r_exponent <- -0.25             # theoretical scaling.
r_scale <- 0.3                  # estimate of r from Daphnia pulex in Table 1 of Fenchel 1974. Original study is Frank et al. 1957       
r_intercept <- log(r_scale) - r_exponent * log(g_zoop) # scale based on zooplankton size

r_benthos_estimate <- mass_scaling(W = g_benthic_invert, intercept = r_intercept, exponent = r_exponent)

#### CARRYING CAPACITY (K) ----
# proportional to M^(-3/4) from Rip and McCann 2011 (they cite Enquist et al. 1998 and Brown et al. 2004)
# alternatively use empirical exponent = -0.77 from Brown et al. 2004
K_exponent <- -0.75 
K_scale <- 5          # admittedly, this is a bit of an arbitrary number. It does correspond to zooplankton biomass density (g/500 L) from fishless ponds
K_intercept <- log(K_scale) - K_exponent * log(g_zoop) # scale based on zooplankton size

K_benthos_estimate <- mass_scaling(W = g_benthic_invert, intercept = K_intercept, exponent = K_exponent)

#### CONVERSION EFFICIENCY (e) ----
# from Mittelbach 1981 (see refs to Ware 1975 and Elliot 1976) 
# independent of body size (Rip and McCann 2011, Yodzis & Innes 1992)
e <- 0.7

#### HABITAT PREFERENCE (w) ----
# Based on % littoral Carbon from Matthews et al. 2010
limnetics_littoral <- c(0.4, 0.1)
benthics_littoral <- c(0.7, 0.6)
solitary_littoral <- c(0.5, 0.6, 1.0, 0.1)

w_limnetics_in_open_water <- 1 - mean(limnetics_littoral) # 0.65 is approximated from Fig. 2 in Schluter 1993
w_benthics_in_benthos <- mean(benthics_littoral) # 0.9 is approximated from Fig. 2 in Schluter 1993

# as 1 - littoral to correspond to parameters in ECD_model.C1.3sp 
w_solitary <- 1 - mean(solitary_littoral)                 # 0.5

#### MORTALITY RATE (m) ----
# m = individual deaths per unit time
# proportional to M^(-1/4) from Brown et al. 2004
# empirical estimate of exponent = -0.24 from Fig. 4 in Brown et al. 2004

# estimate of annual threespine stickleback mortality from fishbase
library(rfishbase) 
threespine_stickleback_M <- filter(popgrowth(species_list = "Gasterosteus aculeatus"), M != "NA")$M

m_exponent <- -0.25                   # theoretical exponent
m_scale <- threespine_stickleback_M        
#m_intercept <- log(m_scale) - m_exponent * log(g_solitary) # use solitary stickleback mass for scaling

m_limnetic_estimate <- m_scale #mass_scaling(W = g_limnetic, intercept = m_intercept, exponent = m_exponent); m_limnetic_estimate
m_benthic_estimate <- m_scale #mass_scaling(W = g_benthic, intercept =  m_intercept, exponent = m_exponent); m_benthic_estimate

#### HANDLING TIME (h) ---

# from Mittelbach 1981, Table 3: Minimum handling time for Daphnia vs. Chironomus. Units = seconds per individual
min_h_Chironomus <- 9.63  # this can be much higher because stickleback than sunfish
min_h_Daphnia <- 1.02
benthos_relative_handling_time <- min_h_Chironomus/min_h_Daphnia

h_scale <- 0.01 # assuming it's independent of consumer body size. This assumption is likely okay, given that prey varied 2-orders of magnitude while stickleback vary much less than 1-order.

#### ATTACK RATE (a) ----
# estimates from Fig. 4, Schluter 1993
# although these are estimates of prey volume consumed per strike, 
# Schluter noted that these relative estimates were virtually identical to intake rate.
# The intake rate data taken by Schluter  is more in line with per capita attack rate; however,
# not all of these data were reported. This is why we are using capture success as a proxy.
# We use these estimates instead of mass-based estimates of attack rate (e.g. Rip and McCann 2011 cite Peters 1983 and Brown et al. 2004),
# since these traits depend other evolved traits that are not related to body size (e.g. gill rakers, gape size and body shape)

limnetic_capture_success_open_water <- 0.011
limnetic_capture_success_benthos <- 0.025        # note that limnetics have an overall higher capture succes in benthos vs. open water, this doesn't nullify the existence of a trade-off between the two habitats
benthic_capture_success_open_water <- 0.004 
benthic_capture_success_benthos <- 0.16

solitary_capture_success_open_water <- 0.009
solitary_capture_success_benthos <- 0.04

a_scale <- 300                                   # scaled until I was able to get coexistence for species pairs

#### CONSUMER-RESOURCE DYNAMICS ----

## Attack rate
A <- 5 #2 
aii <- A*0.6
aij <- eval(a12_tradeoff, data.frame(A=A, a11m=aii, n=n_linear)) # n_concave

m_solitary_McCann <- expression(e11*(((w11m * R1)/(w11m * R1 + w12m * R2)) * a11m * R1)/(1 + ((w11m * R1)/(w11m * R1 + w12m * R2)) * a11m * h11 * R1 + ((w12m * R2)/(w11m * R1 + w12m * R2)) * a12m * h12 * R2) + e12*(((w12m * R2)/(w11m * R1 + w12m * R2)) * a12m * R2)/(1 + ((w11m * R1)/(w11m * R1 + w12 * R2)) * a11m * h11 * R1 + ((w12m * R2)/(w11m * R1 + w12m * R2)) * a12m * h12 * R2) - m1)

w12_tradeoff <- expression(1-w11m)

## Solitary lakes
solitary.init.states <- c(R1=0.1, R2=0.1, C1=0.01) # note that equilibrium dynamics appear to be independent of initial state variables
solitary.parms <- data.frame(r1 = r_scale*g_zoop*100000, 
                             r2 = r_benthos_estimate*g_benthic_invert*100000, 
                             K1 = K_scale*g_zoop*100000, 
                             K2 = K_benthos_estimate*g_benthic_invert*100000,
                             e11 = e, 
                             e12 = e, 
                             #e21 = e, 
                             #e22 = e,
                             h11 = h_scale, 
                             h12 = h_scale*benthos_relative_handling_time, 
                             #h21 = h_scale, 
                             #h22 = h_scale*benthos_relative_handling_time,
                             a11 = aii, #a_scale*solitary_capture_success_open_water, 
                             a12 = aij, #a_scale*solitary_capture_success_benthos, 
                             #a21 = a_scale*benthic_capture_success_open_water, 
                             #a22 = a_scale*benthic_capture_success_benthos,
                             w11 = 0.6, #w_solitary,
                             w12 = 0.4,
                             #w22 = w_benthics_in_benthos, 
                             m1 = m_scale) 
                             #m2 = m_benthic_estimate)

solitary_McCann_concave <- eco_evo_CR(init_parameters = solitary.parms, 
                                          init_states = states_1C_2R, #solitary.init.states, 
                                          eco_CR_model = McCann_1C_2R, 
                                          AD_CR_models = c(m_solitary_McCann),
                                          species_traits = c(1,1), 
                                          eco_param_names = c("a11","w11"), 
                                          eco_tradeoff_names = c("a12","w12"),  
                                          evo_param_names = c("a11m","w11m"),  
                                          evo_tradeoff_names = c("a12m","w12m"), 
                                          tradeoff_functions = c(a12_tradeoff, w12_tradeoff),
                                          extra_tradeoff_params = c(A=A, n=n_concave),
                                      mut.size=0.01)
solitary_McCann_concave[c(1,nrow(solitary_McCann_concave)), ]
## Species-pair lakes
species_pair.init.states <- c(R1=1, R2=1, C1=0.25, C2=0.25) # note that equilibrium dynamics appear to be independent of initial state variables
species_pair.parms <- data.frame(r1 = r_scale, 
                                r2 = r_benthos_estimate, 
                                K1 = K_scale, 
                                K2 = K_benthos_estimate,
                                e11 = e, 
                                e12 = e, 
                                e21 = e, 
                                e22 = e,
                                h11 = h_scale, 
                                h12 = h_scale*benthos_relative_handling_time, 
                                h21 = h_scale, 
                                h22 = h_scale*benthos_relative_handling_time,
                                a11 = a_scale*limnetic_capture_success_open_water, 
                                a12 = a_scale*limnetic_capture_success_benthos, 
                                a21 = a_scale*benthic_capture_success_open_water, 
                                a22 = a_scale*benthic_capture_success_benthos,
                                w11 = w_limnetics_in_open_water, 
                                w22 = w_benthics_in_benthos, 
                                m1 = m_limnetic_estimate, 
                                m2 = m_benthic_estimate)

sim_2C_2R_McCann_concave <- eco_evo_CR(init_parameters = species_pair.parms, 
                                          init_states = species_pair.init.states, 
                                          eco_CR_model = McCann_2C_2R, 
                                          AD_CR_models = c(mC1_McCann, mC2_McCann),
                                          species_traits = c(1,2), 
                                          eco_param_names = c("a11","a22"), 
                                          eco_tradeoff_names = c("a12","a21"),  
                                          evo_param_names = c("a11m","a22m"),  
                                          evo_tradeoff_names = c("a12m","a21m"), 
                                          tradeoff_functions = c(a12_tradeoff, a21_tradeoff),
                                          extra_tradeoff_params = c(A=A, n=n_concave))

@@

## Effect of character divergence on food-web stability

@@
